Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-21 paper/paper.tex
non-linear figure work in progress

    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma N X}{\ctxwffjdg \Gamma {M \equiv N}}

@@

@@

    \inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg
      \Gamma {A \equiv B}}
  \end{mathpar}
  \caption{Formation rules}
  \label{fig:formation}
\end{figure}

\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\unquoteTy X} U}

@@

We give an overview of the linear types and terms in
\Cref{fig:linear-types-and-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and included in the
appendix. First, the linear variable rule says that a linear variable
can be used if it is the \emph{only} variable in the context.

@@

\begin{figure}
  \begin{footnotesize}
  \begin{mathpar}
    \boxed{\linterm \Gamma \Delta {e} {A}}

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : A \lto B \\ \Gamma ; \Delta' \vdash e' : A} {\Gamma ; \Delta, \Delta' \vdash \applto {e} {e'} : B}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \tol A \\ \Gamma ; \Delta' \vdash e' : A}{\Gamma ; \Delta', \Delta \vdash \apptol e {e'} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}    
  \end{footnotesize}
  \caption{Linear terms}
  \label{fig:linear-types-and-terms}
\end{figure}

>>>>

    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}

    \inferrule{\nonlinterm \Gamma {X \equiv Y} {U}}{\ctxwffjdg \Gamma {X \equiv Y}}

@@

    \inferrule{c \in \Sigma}{\linctxwffjdg \Gamma {\literal{c}}}

@@

    \boxed{\linctxwffjdg \Gamma {A \equiv B}}

    \inferrule{\nonlinterm \Gamma {A \equiv B} L}{\linctxwffjdg
      \Gamma {A \equiv B}}

    \inferrule{\lintctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
        {\quoteTy A} \equiv A}}
  \end{mathpar}
  \caption{Formation rules}
  \label{fig:formation}
\end{figure}

\steven{need to write the rules for the equalities of codes
  $M \equiv \quoteTy {\unquoteTy M} : U$ in an equational theory figure for
  non-linear stuff? Same for the codes in universe $L$.

  I will put the nonlinear equational rules, which are so
  unnecessary, in the appendix}
\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

    \inferrule{~}{\nonlinterm {\Gamma , x : X , \Gamma'} {x} {X}}

    \inferrule{\nonlinterm \Gamma M Y \and \ctxwffjdg \Gamma {X \equiv Y}}{\nonlinterm {\Gamma} {M} {X}}

    \inferrule{~}{\nonlinterm {\Gamma} {tt} {1}}

    \inferrule{\nonlinterm \Gamma M X \and \nonlinterm \Gamma {N} {\subst Y {M}
        {x}}}{\nonlinterm {\Gamma} {(M , N)} {\SigTyLimit {x} {X} {Y}}}

    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
              {\nonlinterm {\Gamma} {M.fst} X}

    \inferrule{\nonlinterm {\Gamma} {M} {\SigTyLimit {x} {X} {Y}}}
              {\nonlinterm {\Gamma} {M.snd} {\subst Y {M.fst} {x}}}

    \inferrule{\ctxwffjdg \Gamma {\PiTyLimit {x} {X} {Y}} \and \nonlinterm
      {\Gamma , x : X} {M} {Y}}
              {\nonlinterm {\Gamma} {\lamb {x} {M}} {\PiTyLimit {x} {X} {Y}}}

    \inferrule{\nonlinterm {\Gamma} {M} {\PiTyLimit {x} {X} {Y}} \and
      \nonlinterm \Gamma N X}
              {\nonlinterm \Gamma {M~N} {\subst Y {N} {x}}}

    \inferrule{\nonlinterm \Gamma {M \equiv N} X}{\nonlinterm {\Gamma} {refl} {M
      =_{X} N}}

    \inferrule{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma {\quoteTy X} U}

@@

\steven{todo}
We give an overview of the linear types and terms in
\Cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and included in the
appendix. First, the linear variable rule says that a linear variable
can be used if it is the \emph{only} variable in the context.

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \boxed{\linterm \Gamma \Delta {e} {A}}

    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\nonlinterm \Gamma {M} {\ltonl A}}{\linterm {\Gamma} {\cdot} {M} {A}}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : A \lto B \\ \Gamma ; \Delta' \vdash e' : A} {\Gamma ; \Delta, \Delta' \vdash \applto {e} {e'} : B}
    \\
    %
    \inferrule{\Gamma ; a : A , \Delta \vdash e : B}{\Gamma ; \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e'
      : B \tol A}{\Gamma ; \Delta, \Delta' \vdash \apptol {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \\
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}
  \caption{Linear terms}
  \label{fig:linear-terms}
\end{figure}
